---
title: "project on reproducible research"
author: "sconfumberto"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Exploratory analysis of activity monitoring data

In this R Markdown document is described the step-by-step process to accomplish the 4 tasks required by the project.

### Loading packages and the data
In the code chunk below, the csv file is extraced and read into R

```{r}

#loads the required packages
library(ggplot2)
library(knitr)
library(readr)
library(dplyr)

#extracts the csv file from the zip 
url <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"

working_dir <- getwd()

dest_zip_file <- file.path(working_dir, "data.zip")

dest_csv_file <- file.path(working_dir, "activity.csv")

download.file(url, dest_zip_file, method = "auto")

unzip(dest_zip_file, files = "activity.csv", exdir = working_dir)

if(file.exists(dest_csv_file)) {
  cat("CSV file extracted successfully")
} else {
  cat("Failed to extract the CSV file")
}

#reads the csv file
data <- read_csv("activity.csv")
```

### Preliminary cleaning of the dataset
The first assignment (below) suggested that I could ignore the missing values (NAs), however I have interpreted this as a way of saying that I could ignore them, but also that I could not if I wanted to. As a consequence, I decided to deal with missing values right away. Therefore, task 3 below, will not deal anymore with the issue of imputing missing values.

Furthermore, in this task, I have also dealt right away with the days of the week, because I thought it was important to know from the beginning which days were weekday and which ones were weekends. Therefore, task 4 below, will not deal anymore with this issue of finding the day of the week.

```{r}
#counts NAs in each column
na_count_steps <- sum(is.na(data$steps))
print(na_count_steps)

na_count_date <- sum(is.na(data$date))
print(na_count_date)

na_count_interval <- sum(is.na(data$interval))
print(na_count_interval)

#only the steps column has NAs

#adds a column with the days of the week
# Add a new column indicating weekday or weekend
data$day_name <- ifelse(weekdays(data$date) %in% c("Saturday", "Sunday"), "Weekend", "Weekday")

#calculates the median and mean of steps on both weekdays and weekends
median_weekdays <- data %>%
  filter (day_name == "Weekday") %>%
  summarize(med_days = median(steps, na.rm = TRUE))

median_weekends <- data %>%
  filter (day_name == "Weekend") %>%
  summarize(med_ends = median(steps, na.rm = TRUE))

mean_weekdays <- data %>%
  filter (day_name == "Weekday") %>%
  summarize(mean_days = mean(steps, na.rm = TRUE))

mean_weekends <- data %>%
  filter (day_name == "Weekend") %>%
  summarize(mean_ends = mean(steps, na.rm = TRUE))

#since the median of both weekdays and weekends is the same, i.e. 0, I substitute 0 for NAs in the whole steps column
data$steps[is.na(data$steps)] <- 0
```

### What is mean total number of steps taken per day?
See the code chunk below

```{r}
total_steps <- data %>%
  group_by(date) %>%
  summarize (total_per_day = sum(steps))

#prints the total steps
print (total_steps)

mean_per_day <- mean(total_steps$total_per_day)

#prints the mean
print (mean_per_day)

median_per_day <- median(total_steps$total_per_day)

#prints the median
print (median_per_day)

#makes the plot
hist(total_steps$total_per_day, 
     main = "Histogram of Total Steps per Day",
     xlab = "Total Steps",
     ylab = "Frequency",
     col = "blue")
```

### What is the average daily activity pattern?
See the code chunk below

```{r}
#aggregates the data and calculate the average number of steps per 5-minute interval
aggregated <- aggregate(steps ~ interval, data = data, mean)

plot(aggregated$interval, aggregated$steps,
     type = "l",
     main = "Average Number of Steps by 5-Minute Intervals",
     xlab = "5-Minute Interval",
     ylab = "Average Number of Steps")

#finds the 5-minute interval with the maximum average number of steps
max_interval <- data$interval[which.max(data$steps)]

#prints the result
print(paste("The 5-minute interval with the maximum average number of steps is:", max_interval))

#finds the day with the maximum average number of steps for a specific interval
max_steps_interval <- 615  # Specify the specific 5-minute interval
subset_data <- subset(data, interval == max_steps_interval)
max_day <- subset_data$date[which.max(subset_data$steps)]

#prints the result
print(paste("On", max_day, "the", max_steps_interval, "interval has the maximum average number of steps."))
```

### Imputing missing values
As I have explained above, I have decided to deal with missing value in the cleaning phase, before starting the exploratory analysis.
I have substituted NAs with 0 as this is the value corresponsing to the median across the entire dataset, both for weekdays and weekends.
As a consequence, the dataset without missing values is the imported dataset after the cleaning phase described in the first code chunk above.
In this task, I have only dealt with the last part of the assignment, i.e. to make a histogram of the total number of steps taken each day and calculate and report the mean and median total number of steps taken per day.

Because of the way I have decided to impute missing values, the resulting histogram was heavily left-skewed toward the zero. For this reason, I decided to perform an additional analysis without the interval in which there are 0 steps, to see how the distribution changes, both concerning mean and median values.

```{r}
#makes a histogram of the total number of steps taken each day
hist(data$steps, breaks = "FD", main = "Histogram of Total Steps Per Day",
     xlab = "Total Steps", ylab = "Frequency")

#calculates and reports the mean and median total number of steps taken per day
mean_steps <- mean(data$steps)
median_steps <- median(data$steps)

print(paste("Mean steps per day:", mean_steps))
print(paste("Median steps per day:", median_steps))

#filters out the 0 values from the 'steps' column
non_zero_steps <- data$steps[data$steps != 0]

#makes a histogram of the non-zero values
hist(non_zero_steps, breaks = "Sturges", main = "Histogram of Non-Zero Steps Per Day",
     xlab = "Total Steps", ylab = "Frequency")

#laculates and reports the mean and median total number of steps taken per day
mean_steps <- mean(non_zero_steps)
median_steps <- median(non_zero_steps)

print(paste("Mean steps per day:", mean_steps))
print(paste("Median steps per day:", median_steps))
```

### Are there differences in activity patterns between weekdays and weekends?
See the code chunk below, there are significant differences in activity patterns between the two types of days. This can be clearly seen in the panel plot below

```{r}
data$day_name <- factor(data$day_name, levels = c("Weekday", "Weekend"))

#calculates the average number of steps per interval and day type
avg_steps <- data %>%
  group_by(interval, day_name) %>%
  summarize(avg_steps = mean(steps)) %>%
  ungroup ()

#creates the panel plot
ggplot(avg_steps, aes(x = interval, y = avg_steps, group = day_name, color = day_name)) +
  geom_line() +
  labs(x = "5-Minute Interval", y = "Average Number of Steps",
       title = "Average Number of Steps by Interval") +
  theme_minimal() +
  facet_grid(. ~ day_name)
```
